---
title: "Backtest"
subtitle: "Honest out-of-sample performance"
date: today
date-format: "D MMMM YYYY"
---

```{r setup}
#| include: false
files <- list.files("R", pattern = "\\.R$", recursive = TRUE, full.names = TRUE)
for (f in files) sys.source(f, envir = globalenv())
library(dplyr); library(ggplot2); library(gt); library(scales)

scored_path <- "data/backtest_scored.rds"
meta_path   <- "data/backtest_meta.rds"
scored <- if (file.exists(scored_path)) readRDS(scored_path) else NULL
meta   <- if (file.exists(meta_path))   readRDS(meta_path)   else NULL
has_backtest <- !is.null(scored) && nrow(scored) > 0L
```

## Methodology

For each historical as-of date $T$ on a quarterly grid:

1. Restrict every input series to observations whose publication date is
   $\leq T$ (using documented publication lags — OAD ~35 days, NOM ~183
   days, visa grants ~45 days).
2. Refit every model from scratch on this restricted panel.
3. Score predictions at horizons $h \in \{-2, -1, 0, +1\}$ quarters
   relative to $T$, where $h = 0$ is the "current" quarter (NOM not yet
   published) and $h = -2, -1$ are recent backcasts.
4. Use the latest available ABS NOM as the evaluation "truth".

This is **pseudo-real-time** — we don't have a true vintage history,
so the underlying data values themselves don't reflect their
contemporaneous releases. A real vintage-aware backtest is a logged
follow-up; see [STATUS.md](https://github.com/DavidAStephan/NOM_Nowcast/blob/main/STATUS.md).

Models compared:

| Model | What it uses |
|-------|--------------|
| `bayes_headline` | **Phase 3 v0.1.** Same trivariate observation structure as `kalman_multi_v2` but fit via HMC in Stan, so the credible intervals come from proper posterior sampling instead of treating MLE parameters as fixed. Headline-only. Only fit on the last `cfg$models$bayes_headline$backtest_window` (default 12) quarters of the backtest grid. |
| `bayes_gamma`    | **Phase 3 v0.2.** Extends `bayes_headline` with a Gamma-weighted forward lag from visa grants to arrivals (matches `kalman_v4`'s parametric Gamma lag). The visa observation row loads on `log_sum_exp` of future latent log-arrivals weighted by a fixed Gamma(α, β) pmf. Backtest window same as `bayes_headline`. |
| `kalman_multi_v2` | Multivariate Kalman: OAD + visa grants + NOM as joint observations on a shared latent log-arrivals state. Phase 2 v4.0 with Gamma-lag state augmentation. |
| `kalman_multi_pi` | Same multi-SSM fit, but headline assembled via the empirical-π route instead of the direct NOM observation. |
| `kalman_v1` | Phase 1 univariate Kalman per category, π-based headline. |
| `bridge` | OLS bridge regression: NOM ~ lagged OAD + lagged visa grants. |
| `random_walk` | Headline NOM at $T-1$ rolled forward. |
| `ar1` | AR(1) on quarter-on-quarter NOM growth. |
| `abs_preliminary` | The ABS NOM value at vintage $T$ (informational — we don't have true vintage history). |

The backtest is computed offline (via `scripts/refresh_backtest.R`)
and the scored tibble is committed to the repo. The site renders cheaply
on every push; the backtest refreshes on demand or on a slower cadence.

## Headline results

```{r headline-table}
#| eval: !expr has_backtest
scored |>
  dplyr::filter(category == "total") |>
  dplyr::group_by(model, horizon) |>
  dplyr::summarise(
    n    = sum(n,   na.rm = TRUE),
    RMSE = mean(rmse, na.rm = TRUE),
    MAE  = mean(mae,  na.rm = TRUE),
    Bias = mean(bias, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::filter(is.finite(RMSE)) |>
  dplyr::arrange(horizon, RMSE) |>
  dplyr::mutate(Horizon = paste0("h = ", horizon)) |>
  dplyr::select(Horizon, Model = model, n, RMSE, MAE, Bias) |>
  gt::gt(groupname_col = "Horizon") |>
  gt::fmt_number(columns = c("RMSE", "MAE", "Bias"),
                 decimals = 0, use_seps = TRUE) |>
  gt::tab_options(table.font.size = "small") |>
  gt::tab_source_note(gt::md(
    paste0(
      "Backtest grid: ",
      if (!is.null(meta)) meta$n_successful else "?",
      " of ",
      if (!is.null(meta)) meta$n_asof_dates else "?",
      " quarterly as-of dates",
      if (!is.null(meta)) paste0(" from ",
            nn_quarter_label(meta$grid_from, sep = "-"), " to ",
            nn_quarter_label(meta$grid_to,   sep = "-"))
      else "",
      ". Refreshed ",
      if (!is.null(meta)) format(meta$refreshed_at, "%d %B %Y") else "—",
      ". RMSE / MAE in persons. Lower is better. ",
      "Models are sorted within each horizon by RMSE."
    )
  ))
```

```{r placeholder}
#| eval: !expr (!has_backtest)
#| results: asis
cat(
  ":::{.callout-note}\n",
  "**Backtest snapshot not yet committed to this repo.** Run ",
  "`Rscript scripts/refresh_backtest.R` locally, or trigger the ",
  "`Refresh backtest` workflow on GitHub. The script writes ",
  "`data/backtest_scored.rds`, which this page reads on every render.\n",
  ":::\n",
  sep = ""
)
```

## RMSE by horizon

```{r rmse-plot}
#| eval: !expr has_backtest
plot_backtest_rmse(scored)
```

## Notes on interpretation

- **Bridge regression dominates backcasts** (h = -2, -1) because it
  regresses NOM directly on lagged flows with no log-Gaussian
  assumption baggage. Its h = 0 RMSE looks worst because at the
  current-quarter horizon many of its regressors are missing (OAD
  publication lag) and `predict()` returns NA for those rows. Those
  NAs propagate to a few large errors.
- The **multivariate Kalman** is the recommended production model
  because it gives unified uncertainty quantification across all
  horizons and produces credible intervals on the headline. Bridge
  regression doesn't natively give intervals on out-of-sample
  forecasts.
- **No model has direct NOM signal at h = 0 / h = +1** (NOM isn't
  published yet at those horizons), so all Kalman variants converge
  to similar performance. This is the structural limit of point
  estimation from public data; bigger gains require either richer
  leading indicators or a Bayesian model with explicit posterior
  uncertainty (Phase 3 punch-list item).
- **The ABS preliminary "model"** is informational only — without a
  captured vintage history the framework falls back to using the
  latest available NOM as the preliminary value, which is circular.
  A real vintage-aware backtest would require running the pipeline
  weekly for ~6 months to accumulate ABS revisions.

## Run details

```{r run-meta}
if (!is.null(meta)) {
  cat(sprintf("- Backtest as-of dates: %d (%s to %s)\n",
              meta$n_asof_dates,
              as.character(meta$grid_from),
              as.character(meta$grid_to)))
  cat(sprintf("- Last refreshed:       %s\n",
              format(meta$refreshed_at, "%d %B %Y %H:%M %Z")))
} else {
  cat("- No backtest snapshot present yet.\n")
}
cat(sprintf("- Site build:           %s\n",
            format(Sys.time(), "%d %B %Y %H:%M %Z")))
```
