---
title: "Australian NOM Nowcast"
subtitle: "Real-time estimate of quarterly Net Overseas Migration"
date: today
date-format: "D MMMM YYYY"
---

```{r setup}
#| include: false
files <- list.files("R", pattern = "\\.R$", recursive = TRUE, full.names = TRUE)
for (f in files) sys.source(f, envir = globalenv())
library(dplyr); library(ggplot2); library(gt); library(scales); library(lubridate)

cfg   <- config::get(file = "config.yml")
asof  <- Sys.Date()

# All-in-one site build: fetch + fit. Targets cache is reused when
# available so this should be cheap on repeat renders.
panel <- tryCatch(targets::tar_read(panel_quarterly), error = function(e) NULL)
if (is.null(panel) || !nrow(panel)) {
  oad <- fetch_oad(cfg, asof); nom <- fetch_nom(cfg, asof); vg <- fetch_visa_grants(cfg, asof)
  panel <- build_quarterly_panel(clean_oad(oad, cfg), clean_nom(nom, cfg),
                                 clean_visa_grants(vg, cfg), empty_students_clean(), cfg)
}
fit_multi <- fit_kalman_multi(panel, cfg)
fc_multi  <- forecast_kalman_multi(fit_multi, asof, cfg)
fits_uni  <- fit_kalman_univariate(panel, cfg)
fc_uni    <- forecast_kalman_univariate(fits_uni, asof, cfg)
pi_emp    <- estimate_pi_empirical(panel, cfg)
pi_sm     <- smooth_pi(pi_emp, cfg)

headline <- build_nowcast_headline_from_multi(fc_multi, cfg)

# Identify the current quarter and the most recent ABS-published quarter
asof_q  <- nn_quarter_start(asof - lubridate::days(1L))
abs_obs <- panel |>
  dplyr::filter(category == "total", !is.na(nom_final)) |>
  dplyr::arrange(period)
latest_abs <- if (nrow(abs_obs)) tail(abs_obs, 1L) else NULL

# Pick the headline row: the latest forecast quarter for which ABS hasn't yet
# published a number.
hl_row <- headline |>
  dplyr::filter(period > (if (!is.null(latest_abs)) latest_abs$period else as.Date("1900-01-01"))) |>
  dplyr::arrange(period) |>
  dplyr::slice(1L)

fmt_k <- function(x) {
  if (is.na(x)) return("—")
  paste0(scales::number(x / 1000, accuracy = 0.1), "k")
}
```

::: {.callout-card}
::: {.headline-label}
Headline nowcast — `{r} nn_quarter_label(hl_row$period, sep = ' · ')`
:::
::: {.headline-number}
`{r} fmt_k(hl_row$nom_mean)`
:::
::: {.headline-interval}
80% interval: `{r} fmt_k(hl_row$lower_80)` to `{r} fmt_k(hl_row$upper_80)`
:::
::: {.headline-asof}
Quarterly Net Overseas Migration. Run as of `{r} format(asof, '%d %B %Y')`.
:::
:::

## What this is

This site publishes a real-time estimate of Australian quarterly **Net
Overseas Migration (NOM)** built from publicly available data only — Overseas
Arrivals and Departures and quarterly population components from the ABS, and
Department of Home Affairs visa-grant statistics via data.gov.au.

The headline number above is the model's nowcast for the quarter ABS hasn't
yet published. The model is a multivariate state-space Kalman filter
(Phase 2 v4.0) — see [Methodology](methodology.qmd) for the equations and
[Backtest](backtest.qmd) for honest performance numbers.

## Recent quarters

```{r recent-table}
recent <- headline |>
  dplyr::left_join(
    panel |> dplyr::filter(category == "total") |>
      dplyr::select(period, abs_nom = nom_final),
    by = "period"
  ) |>
  dplyr::arrange(period) |>
  dplyr::filter(period >= asof_q %m-% months(15L)) |>
  dplyr::transmute(
    Quarter      = nn_quarter_label(period, sep = " · "),
    Model        = nom_mean,
    `80% lower`  = lower_80,
    `80% upper`  = upper_80,
    ABS          = abs_nom,
    `Error vs ABS` = ifelse(is.na(abs_nom), NA, nom_mean - abs_nom)
  )

recent |>
  gt::gt() |>
  gt::fmt_number(columns = c("Model", "80% lower", "80% upper", "ABS"),
                 decimals = 0, use_seps = TRUE) |>
  gt::fmt_number(columns = "Error vs ABS",
                 decimals = 0, use_seps = TRUE, force_sign = TRUE) |>
  gt::sub_missing(missing_text = "—") |>
  gt::tab_style(style = gt::cell_text(weight = "bold"),
                locations = gt::cells_body(rows = is.na(recent$ABS))) |>
  gt::tab_options(table.font.size = "small") |>
  gt::tab_source_note(gt::md(
    "**Model**: multivariate Kalman with NOM, OAD long-term arrivals, and DHA visa grants as joint observations. **ABS**: latest published NOM via `national-state-and-territory-population` cat. 3101.0. Bolded rows are unpublished by ABS at the asof date — those are nowcasts."
  ))
```

## Nowcast vs ABS over time

```{r headline-plot}
plot_nowcast_headline(headline, panel, asof = asof)
```

## Where the model gets its signal

The state-space Kalman jointly observes three series, all loading on a shared
latent log-arrivals state:

1. **OAD long-term arrivals** (monthly, 5-week publication lag) — the
   primary high-frequency signal.
2. **DHA visa grants** (annual financial-year aggregates broadcast to
   quarters via OAD proportions) — a *leading* indicator: visa grants in
   FY-Y predict arrivals in FY-Y and FY-Y+1.
3. **ABS quarterly NOM total** (~6 month publication lag) — direct
   observation of the target itself, used to anchor the latent state for
   historical quarters.

When ABS hasn't yet published NOM for the current quarter, the model uses
OAD and visa-grants observations to estimate the latent state, then maps
back to NOM via a calibrated offset. See [Methodology](methodology.qmd) for
the equations.

## How accurate is it?

Out-of-sample backtest across 11 quarterly as-of dates from 2023-Q1 to
2025-Q3, with each as-of date's data restricted to what would have been
publicly observable then:

| Horizon              | Multivariate Kalman RMSE |
|----------------------|-------------------------:|
| h = -2 (backcast)    | ~36,000 |
| h = -1 (backcast)    | ~38,000 |
| h = 0  (nowcast)     | ~78,000 |
| h = +1 (forecast)    | ~74,000 |

For context, ABS quarterly NOM is typically 50,000 – 150,000 per quarter.
RMSE around 36k on backcasts corresponds to ~25% relative error on a
60,000 quarter, ~15% on a 130,000 quarter.

The model performs best on **backcasts** (revising NOM history) — where ABS
NOM is already observed and pins the state down. Nowcasts (current quarter)
have wider uncertainty because the model has no direct NOM signal yet.
Full breakdown on the [Backtest](backtest.qmd) page.

## Run details

```{r run-meta}
n_obs <- nrow(panel)
periods_span <- range(panel$period, na.rm = TRUE)
gl_w <- fit_multi$visa_gamma_weights

cat(sprintf("- Pipeline run:        %s\n", format(asof, "%d %B %Y, %H:%M %Z")))
cat(sprintf("- Panel coverage:      %s to %s (%d quarterly observations × %d categories)\n",
            nn_quarter_label(periods_span[1L], sep = "-"),
            nn_quarter_label(periods_span[2L], sep = "-"),
            length(unique(panel$period)), length(unique(panel$category))))
cat(sprintf("- Model:               %s\n", fit_multi$kind %||% "kalman_multi"))
cat(sprintf("- Alpha_v offset:      %.3f (log-gap OAD-arrivals vs lagged visa-grants)\n",
            fit_multi$alpha_v))
cat(sprintf("- Alpha_n offset:      %.3f (log-gap OAD-arrivals vs NOM)\n",
            fit_multi$alpha_n))
if (!is.null(gl_w)) {
  cat("- Gamma lag weights:   ", paste(round(gl_w, 3), collapse = ", "), "\n")
}
```
