---
title: "NOM Nowcast — Backtest Report"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    embed-resources: true
    fig-width: 9
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(tidyr)
library(ggplot2)
library(gt)
library(targets)

r_files <- list.files(file.path("..", "R"), pattern = "\\.R$",
                      recursive = TRUE, full.names = TRUE)
for (f in r_files) sys.source(f, envir = globalenv())

scored      <- targets::tar_read(backtest_scored)
backtest_runs <- targets::tar_read(backtest_runs)
cfg         <- targets::tar_read(cfg_target)
```

## Overview

This report scores the Phase 1 Kalman-filter pipeline against three
benchmarks on a vintage-aware recursive backtest. The most important
comparison is against the **ABS preliminary estimate**: that is the
number the user would otherwise be relying on at each evaluation date.

```{r summary-totals}
scored |>
  dplyr::filter(category == "total") |>
  dplyr::group_by(model) |>
  dplyr::summarise(
    n     = sum(n, na.rm = TRUE),
    RMSE  = round(mean(rmse, na.rm = TRUE)),
    MAE   = round(mean(mae,  na.rm = TRUE)),
    Bias  = round(mean(bias, na.rm = TRUE)),
    `RMSE vs prelim` = round(mean(rmse_vs_prelim, na.rm = TRUE)),
    `Hit rate QoQ`   = round(mean(hit_rate_qoq, na.rm = TRUE), 2)
  ) |>
  dplyr::arrange(RMSE) |>
  gt::gt() |>
  gt::tab_header(title = "Aggregate performance across all horizons and regimes")
```

## RMSE by horizon and regime

```{r rmse-plot, fig.height=6}
plot_backtest_rmse(scored)
```

## Performance by regime

```{r regime-table}
scored |>
  dplyr::filter(category == "total", horizon == 0) |>
  dplyr::group_by(regime, model) |>
  dplyr::summarise(
    RMSE = round(mean(rmse, na.rm = TRUE)),
    `RMSE vs prelim` = round(mean(rmse_vs_prelim, na.rm = TRUE)),
    .groups = "drop"
  ) |>
  tidyr::pivot_wider(names_from = model,
                     values_from = c(RMSE, `RMSE vs prelim`)) |>
  gt::gt() |>
  gt::tab_header(title = "Nowcast (h = 0) RMSE by regime")
```

## Per-category performance

```{r cat-table}
scored |>
  dplyr::filter(horizon == 0) |>
  dplyr::filter(category != "total") |>
  dplyr::group_by(category, model) |>
  dplyr::summarise(RMSE = round(mean(rmse, na.rm = TRUE)),
                   .groups = "drop") |>
  tidyr::pivot_wider(names_from = model, values_from = RMSE) |>
  gt::gt() |>
  gt::tab_header(title = "Nowcast RMSE by category")
```

## Discussion template

The skeleton report writes mechanical tables. A human-written discussion
section belongs here at each release covering:

- where the structural model adds value over ABS preliminary
- where it does not (and why)
- per-category strengths and weaknesses (typically student is the
  biggest mover; bridging is the hardest)
- robustness across regimes

See [methodology.html](methodology.html) for the underlying
specification.
