---
title: "NOM Nowcast"
subtitle: "Current vintage"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    embed-resources: true
    fig-width: 9
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(tidyr)
library(ggplot2)
library(gt)
library(targets)

# Make project helper functions available in the rendering session.
r_files <- list.files(file.path("..", "R"), pattern = "\\.R$",
                      recursive = TRUE, full.names = TRUE)
for (f in r_files) sys.source(f, envir = globalenv())

cfg            <- targets::tar_read(cfg_target)
asof           <- targets::tar_read(asof_date)
headline       <- targets::tar_read(nowcast_headline)
categories     <- targets::tar_read(nowcast_categories)
panel          <- targets::tar_read(panel_quarterly)
pi_smoothed    <- targets::tar_read(pi_smoothed)
pi_empirical   <- targets::tar_read(pi_empirical)
```

## Headline

```{r headline-table}
latest <- headline |>
  dplyr::arrange(desc(period)) |>
  head(4) |>
  dplyr::transmute(
    Quarter = format(period, "%Y Q%q"),
    `Model NOM`     = scales::comma(round(nom_mean)),
    `80% interval`  = sprintf("[%s, %s]",
                              scales::comma(round(lower_80)),
                              scales::comma(round(upper_80))),
    `95% interval`  = sprintf("[%s, %s]",
                              scales::comma(round(lower_95)),
                              scales::comma(round(upper_95)))
  )
gt::gt(latest) |>
  gt::tab_header(
    title = "Model NOM — most recent four quarters",
    subtitle = paste("As of", format(asof, cfg$reporting$asof_label_fmt))
  )
```

```{r headline-plot}
plot_nowcast_headline(headline, panel)
```

## Category decomposition

```{r category-plot}
plot_nowcast_categories(categories)
```

```{r category-table}
cat_tbl <- categories |>
  dplyr::filter(period == max(period)) |>
  dplyr::transmute(
    Category   = category,
    `Model NOM` = scales::comma(round(nom_mean)),
    `± 80%`     = scales::comma(round(1.282 * nom_se))
  )
gt::gt(cat_tbl) |>
  gt::tab_header(title = "Latest-quarter contribution by category")
```

## π — classification probabilities

```{r pi-plot, fig.height=6}
plot_pi(pi_smoothed, pi_empirical)
```

## Methodology and limitations

See [methodology.html](methodology.html) for the full write-up and
[backtest_report.html](backtest_report.html) for performance evaluation.

Quick recap:

- **Approach.** Per-category univariate Kalman filter on log long-term
  OAD flows; smoothed empirical π applied to convert latent flows to
  NOM contributions.
- **Inputs at this vintage.** ABS OAD (cat. 3401.0), ABS NOM
  (cat. 3101.0). Phase 2 multi-source extensions (DHA visa grants,
  Department of Education enrolments) are scaffolded but not yet active
  in the headline path — they are surfaced as side-by-side diagnostics
  only.
- **Uncertainty.** Credible intervals are model-internal; they do not
  account for π misspecification at regime breaks. Treat intervals as
  conservative lower bounds on true uncertainty during 2020–22.
